Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing

### 2013-04-26 ace/ace.tex
Updated high-level goals

 We begin in Section \ref{usage} with simple examples that show how Ace can be used for low-level GPU programming and introduce the fundamental decisions made in its design. We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and type inference scheme, and an explicit phase separation between compile-time and run-time logic.

The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as metalanguage objects. We refer to such types as \emph{active types} by relation to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}). Unlike prior approaches to extensibility that take a language-global view, which hinders interoperability, the AT\&T mechanism guarantees that extensions are composable by construction by limiting their scope to expressions having a type defined by that extension.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism. First-class support for each of these has required a new language in the past. AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be  used naturally within a single program or workflow.

Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to  integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal overhead. Achieving this deep level of integration makes use of Ace's representation of types as well.

@@

\section{Language Design and Usage}\label{usage}
\subsection{High-Level Design}
Static type systems are powerful tools for programming language design and implementation. By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs. This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions. Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples). Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.

It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC. One reason has to do with the criteria of \textbf{ease-of-use}. Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit type annotations on every function and variable declaration,  separate header files, and other declarations that exist to assist the compiler. Cordy identifies the principle of \emph{conciseness} with elimination of
redundancy and the availability of reasonable defaults \cite{cordy1992hints}. The high-level languages used in HPC have, consistent with this principle, avoided most of this overhead. Ace was originally conceived to explore the question: does this require run-time mechanisms, or can one express a statically-typed, low-level language at least as powerful as C without the overhead that is typically associated with it?

Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, without modification, an existing language, Python's, syntax. This was not an arbitrary choice but rather a key mechanism by which Ace achieves \textbf{continuity}. Python is amongst the most widely-adopted languages in the world, and enjoys similar  ubiquity within HPC specifically \cite{pythonhpc}. By directly adopting Python's syntax, Ace's syntax is immediately familiar and acceptable to a significant portion of it's intended audience. Moreover, any tools that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used without modification for handling Ace code.

\subsection{Example 1: Hello, World!}
\begin{codelisting}
\lstinputlisting{hello.py}
\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
\label{hello}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec hello.py`
Hello, compile-time world!
Goodbye, compile-time world!
$ `cat hello.cl`
__kernel void main() {
    printf("Hello, run-time world.");
}\end{lstlisting}
\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
\label{helloout}
\end{codelisting}

>>>>

 We begin in Section \ref{usage} with simple examples that show how Ace can be used for low-level GPU programming and introduce the fundamental decisions made in its design. We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and inference scheme, and an explicit phase separation between compile-time and run-time logic. Each of these can be seen in Listings 1 and 2, which we will discuss further below.

All examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as first-class objects at compile-time. We refer to these objects as \emph{active types} by analogy with \emph{active libraries} \cite{activelibraries} (see Section \ref{related}). Unlike prior approaches to extensibility where users globally modify the grammar or semantics of a language, and thus introduce conflicts between extensions, AT\&T guarantees that extensions are composable by construction by pairing new rules with new types and limiting their scope to expressions of that type.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism. First-class support for each of these has required a new language in the past. AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be coexist naturally within a single program or workflow.

Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to  integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal overhead. Achieving this deep level of integration makes use of Ace's novel type representation as well.

@@

\section{Language Design and Usage}\label{usage}
\begin{codelisting}
\lstinputlisting{hello.py}
\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
\label{hello}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec hello.py`
Hello, compile-time world!
Goodbye, compile-time world!
$ `cat hello.cl`
__kernel void main() {
    char* hello = "Hello, run-time world!";
    printf(hello);
}\end{lstlisting}
\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
\label{helloout}
\end{codelisting}
\subsection{High-Level Design}
Static type systems are powerful tools for programming language design and implementation. By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs. This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions. Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples). Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.

It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC. Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}. More specifically, Cordy identifies the principle of \emph{conciseness} with elimination of
redundancy and the availability of reasonable defaults \cite{cordy1992hints}. Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit and often redundant type annotations on each function and variable declaration,  the use of separate header files, explicit template instantiation and other annotations.  The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system. Ace was originally conceived to explore the question: does conciseness require run-time mechanisms, or can one design a statically-typed language with the same low-level semantics of C with syntactic overhead comparable to a high-level scripting language?

Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, \emph{without modification}, an existing syntax, Python's. This was not an arbitrary choice but rather a key decision by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}. Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{pythonhpc}. By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience. Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used directly without modification as well.

\subsection{Example: Two-Phase ``Hello, World!''}
Ace's \textbf{continuity} with Python does not stop at its syntax.

A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL code is demonstrated in Listing \ref{helloout}.
